Interlocks Mitigate Information Cascades in Large-Scale Social Learning

### 2024-10-30 manuscript.tex
Updates from Overleaf

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Edward L. Platt}
%\authornote{Both authors contributed equally to this research.}
\email{ed@elplatt.com}
\orcid{0000-0003-2148-3841}
\affiliation{%
  \institution{Cinder Oak}
  \city{Ann Arbor}
  \state{Michigan}
  \country{USA}
}
\affiliation{%
  \institution{University of Michigan}
  \city{Ann Arbor}
  \state{Michigan}
  \country{USA}
}

\author{Herminio Bodon}
\orcid{0009-0004-3270-3621}
\email{HerminioBodon2020@u.northwestern.edu}
\affiliation{%
  \institution{Northwestern University}
  \city{Evanston}
  \state{Illinois}
  \country{USA}
}
\affiliation{%
  \institution{University of Michigan}
  \city{Ann Arbor}
  \state{Michigan}
  \country{USA}
}

\author{Daniel M. Romero}
\orcid{0000-0002-8351-3463}
\affiliation{%
  \institution{University of Michigan}
  \city{Ann Arbor}
  \state{Michigan}
  \country{USA}}
\email{drom@umich.edu}

%%
%% By default, the full list of authors will be used in the page
%% headers. Often, this list is too long, and will overlap
%% other information printed in the page headers. This command allows
%% the author to define a more concise list
%% of authors' names for this purpose.
\renewcommand{\shortauthors}{Platt, Bodon, \& Romero}

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}

\end{abstract}

@@

\received{16 January 2024}

@@

\subsection{Interlocks}
When reviewing these examples, a remarkably common pattern stands out: interlocking network structure.
In these successful examples of large-scale participatory projects,
communication is frequently organized into small participatory groups of peers based on interest, region, function, or simply happenstance.
These groups are then interconnected through individuals who participate in multiple groups.
We reproduce excerpts from several examples here to demonstrate the ubiquity of this pattern.

@@

\subsubsection{Properties of the NK Model}
The real-world applicability of our findings will depend on how well a real-world system agrees with the assumptions of our model.
We enumerate key assumptions and properties below.
\begin{description}
\item[Discrete solution space]
The NK model solution space consists of discrete states.
This space best represents a set of yes/no questions.
However, the NK model can also be applied to any solution space that can be approximated by a discrete set of solutions, for example, by dividing a continuous range into bins.
\item[Large solution space]
There are $2^N$ possible solutions for any particular instance of the NK model.
The number of solutions grows exponentially with the number of bits in $s$.
The NK model is thus well-suited to representing problems with far more potential solutions than could ever be evaluated.
For example, for N=33, there would be more potential solutions than humans alive on Earth (at time of writing).
\item[Rugged solution space]
Often called a ``rugged landscape,'' the NK model has many local maxima and minima,
often described as ``hills'' and ``valleys.''
A simple hill-climbing search by changing one solution bit at a time will lead to a local maximum,
but there is no way to know if a better solution exists, other than searching for it.
This is a common (if frustrating) feature of real-world problems.
\item[Tunably rugged]
In addition to being rugged, the ruggedness of the NK model can be ``tuned'' by changing the $K$ parameter.
For a value of $K=0$, each solution bit can be optimized independently, and there is a single local maximum.
For the maximum value of $K=N-1$,
the landscape is essentially random.
For intermediate values of $K$,
there are a large number of local maxima, but they are surrounded by ``basins of attraction'' which allow those maxima to be found from nearby solutions using hill-climbing.
These intermediate values are typically the ones used to mimic real-world rugged landscapes.
\item[NP-complete]
For $K \geq 3$, the problem of finding the global optimum of the NK model in general is NP-complete \cite{weinberger_np_1996}.
This means that the NK model is computationally equivalent to the hardest problems solvable by a computer,
and can thus model the complexity of any problem that is capable of being modeled computationally.
\end{description}

@@

\subsubsection{Small-World}
In {\em small-world} networks
\cite{watts_collective_1998},
individuals are mostly linked to others in the
same spatial or geographic location,
but have some long-distance {\em shortcuts}.
The local ties create high {\em clustering}:
an individuals neighbors are likely to be neighbors to each other.
The shortcuts act as weak ties \cite{granovetter_strength_1973},
reducing the ``size'' (mean shortest path length)
of the network by creating short paths
between any two individuals.
In Watts-Strogatz small-world network \cite{watts_collective_1998},
individuals are assigned to locations on a ring and connected to their
$k$ nearest neighbors (not to be confused with $K$ from the NK model).
For each of an individuals links,
with a probability $\beta$,
the other end of the link is assigned to a random individual,
creating the shortcut edges.

\subsubsection{Preferential-Attachment}

\subsubsection{Interlocks}

@@

Agent-based model parameters

Random-Pod

@@

\section{Discussion}

%%
%% The acknowledgments section is defined using the "acks" environment
%% (and NOT an unnumbered section). This ensures the proper
%% identification of the section in the article metadata, and the
%% consistent spelling of the heading.
\begin{acks}

\end{acks}

@@

\section{Research Methods}

\section{Online Resources}

>>>>

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
%\authornote{Both authors contributed equally to this research.}
\author{Authors redacted for peer review}
%\author{Edward L. Platt}
%\email{ed@elplatt.com}
%\orcid{0000-0003-2148-3841}
%\affiliation{%
%  \institution{Cinder Oak}
%  \city{Ann Arbor}
%  \state{Michigan}
%  \country{USA}
%}
%\affiliation{%
%  \institution{University of Michigan}
%  \city{Ann Arbor}
%  \state{Michigan}
%  \country{USA}
%}

%\author{Herminio Bodon}
%\orcid{0009-0004-3270-3621}
%\email{HerminioBodon2020@u.northwestern.edu}
%\affiliation{%
%  \institution{Northwestern University}
%  \city{Evanston}
%  \state{Illinois}
%  \country{USA}
}
%\affiliation{%
%  \institution{University of Michigan}
%  \city{Ann Arbor}
%  \state{Michigan}
%  \country{USA}
}

%\author{Daniel M. Romero}
%\orcid{0000-0002-8351-3463}
%\affiliation{%
%  \institution{University of Michigan}
%  \city{Ann Arbor}
%  \state{Michigan}
%  \country{USA}}
%\email{drom@umich.edu}

%%
%% By default, the full list of authors will be used in the page
%% headers. Often, this list is too long, and will overlap
%% other information printed in the page headers. This command allows
%% the author to define a more concise list
%% of authors' names for this purpose.
%\renewcommand{\shortauthors}{Platt, Bodon, \& Romero}

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
Large-scale participatory decision-making is notoriously difficult, but successful examples have recently become more common. These examples have been enabled in-part by the fast, affordable, many-to-many communication afforded by the internet. But not all large-scale participatory projects on the internet are successful. We review the empirical literature on successful large-scale participatory projects and identify a common feature: interlocking network structure. We use agent-based modelig to simulate the effects of interlocking network structure on large-scale social learning. We find that in high-conformity contexts, interlocks can provide a protective effect against information cascades and polarization. Our findings suggest that interlocking network structure is beneficial for large-scale participatory projects, and we identify likely mechanisms behind that benefit.
\end{abstract}

@@

\received{29 October 2024}

@@

\subsection{Interlocks\label{subsec:interlocks}}
When reviewing these examples, a remarkably common pattern stands out: interlocking network structure.
In these successful examples of large-scale participatory projects,
communication is frequently organized into small participatory groups of peers based on interest, region, function, or simply happenstance.
These groups are then interconnected through individuals who participate in multiple groups.
We reproduce excerpts from several examples here to demonstrate the ubiquity of this pattern.

@@

\subsubsection{Properties of the NK Model}
The real-world applicability of our findings will depend on how well a real-world system agrees with the assumptions of our model.
We enumerate key assumptions and properties below.
\begin{description}
\item[Discrete solution space]
The NK model solution space consists of discrete states.
This space best represents a set of yes/no questions.
However, the NK model can also be applied to any solution space that can be approximated by a discrete set of solutions, for example, by dividing a continuous range into bins.
\item[Large solution space]
There are $2^N$ possible solutions for any particular instance of the NK model.
The number of solutions grows exponentially with the number of bits in $s$.
The NK model is thus well-suited to representing problems with far more potential solutions than could ever be evaluated.
For example, for N=33, there would be more potential solutions than humans alive on Earth (at time of writing).
\item[Rugged solution space]
Often called a ``rugged landscape,'' the NK model has many local maxima and minima,
often described as ``peaks'' and ``valleys.''
A simple hill-climbing search by changing one solution bit at a time will lead to a local maximum,
but there is no way to know if a better solution exists, other than searching for it.
This is a common (if frustrating) feature of real-world problems.
\item[Tunably rugged]
In addition to being rugged, the ruggedness of the NK model can be ``tuned'' by changing the $K$ parameter.
For a value of $K=0$, each solution bit can be optimized independently, and there is a single local maximum.
For the maximum value of $K=N-1$,
the landscape is essentially random.
For intermediate values of $K$,
there are a large number of local maxima, but they are surrounded by ``basins of attraction'' which allow those maxima to be found from nearby solutions using hill-climbing.
These intermediate values are typically the ones used to mimic real-world rugged landscapes.
\item[NP-complete]
For $K \geq 3$, the problem of finding the global optimum of the NK model in general is NP-complete \cite{weinberger_np_1996}.
This means that the NK model is computationally equivalent to the hardest problems solvable by a computer,
and can thus model the complexity of any problem that is capable of being modeled computationally.
\end{description}

@@

\subsubsection{Small-World}
In {\em small-world} networks
\cite{watts_collective_1998},
individuals are mostly linked to others in the
same spatial or geographic location,
but have some long-distance {\em shortcuts}.
The local ties create high {\em clustering}:
an individuals neighbors are likely to be neighbors to each other.
The shortcuts act as weak ties \cite{granovetter_strength_1973},
reducing the ``size'' (mean shortest path length)
of the network by creating short paths
between any two individuals.
High clustering and short path length are often observed in real-world social
networks.
In a $k$-degree Watts-Strogatz small-world network \cite{watts_collective_1998},
individuals are assigned to locations on a ring and connected to their
$k$ nearest neighbors (not to be confused with $K$ from the NK model).
For each of an individuals links,
with a probability $\beta$,
the other end of the link is assigned to a random individual,
creating the shortcut edges.

\subsubsection{Preferential Attachment}
Some real-world social networks exhibit a skewed, heavy-tail degree distribution:
most individuals linked to few others,
but a few ``hubs'' or ``influencers'' linked to many others.
Such networks can be modeled using the Barabási-Albert preferential attachment model
\cite{barabasi_emergence_1999}.
In this model, as individuals are added to a network,
they are randomly linked to $m$ others.
The probability of such links is proportional to the target's degree
(number of existing links),
creating a ``rich-get-richer'' dynamic and a heavy-tailed degree distribution.
These networks model settings where a small number of individuals produce a disproportionate share of the communication due to factors such as social capital, expertise, or confidence.

@@

\subsection{Modeling Interlocks: Random-Pod Networks\label{subsec:randompod}}
In addition to the two common models of social networks above,
we also model interlock networks,
the central focus of this paper.
Interlock networks are described in detail in Section \ref{subsec:interlocks},

We model an interlock as a time-dependent network.
For a degree $k$ {\em random-pod network},
each agent is randomly assigned to a ``pod''
such that $k+1$ agents are assigned to each pod.
Agents assigned to the same pod are linked by a network edge,
resulting in a fully-connected clique for each pod.
After some number of time-steps, the pod assignments are re-randomized,
creating new cliques.
The movement of agents between pods over the course of the simulation
creates interlocking structure.

By using a time-dependent network, we are able to specify the degree (number of neighbors) of each agent at any given time.
This allows us to choose all of our networks to have the same mean degree,
reducing potential counfounding effects.

The random-pod network model is similar to Salehi and Bernstein's network rotation
\cite{salehi_hive:_2018}.
The key difference is that in network rotation,
pods are changed by a single member at a time, while the random-pod network
reassigns all members simultaneously.
We opt for simultaneous reassignment because it best represents interlock networks
with multiple membership in distinct groups.

\subsection{Agent-Based Model Parameters}
We simulate social-learning using an Agent-Based Model (ABM) closely resembling those of other authors \cite{
lazer_network_2007,
barkoczi_social_2016,
gomez_clustering_2019}.
For our model, we simulate 100 agents.
Each agent maintains a solution candidate and updates that candidate
at each time step according to its learning strategy.
Within each run, all agents use the same learning strategy.
Each run is simulated for 300 time steps
(we have found this is sufficient to achieve convergence for our paremeters).
We created 1000 random instances of the NK model.
For each instance, we simulated each combination of network and learning strategy.

\subsubsection{Objective Function}
Agents search for a binary string $s$ that maximizes an objective
function $Q(s)$ defined by the NK model \cite{kauffman_towards_1987, weinberger_local_1991}.
In keeping with other authors \cite{lazer_network_2007, barkoczi_social_2016}, we use $N=15$, $K=6$.
We also follow the common practice of raising the raw NK fitness to a power of 8
to emphasize peaks and valleys: $Q(s) = Q_{NK}(s)^8$.
Similarly, we normalize fitness values to the global maximum.

\subsubsection{Network}
Agents are placed on a network with the vertices representing
the agents, and edges representing which agents can interact.
We use three separate network topologies:
\begin{description}
    \item[Small-World:]{A 4-degree Watts-Strogatz small-world network with $\beta = 0.1$;}
    \item[Preferential Attachment:]{A Barabási-Albert network with $m=2$ (mean degree 4);}
    \item[Random-Pod:]{A 4-degree random-pod network.}
\end{description}

\subsubsection{Learning Strategy}
At each time step, agents perform individual learning on their previous
solution candidate, followed by one of two social learning strategies:
\begin{description}
\item[Best-Neighbor]{(low-conformity)},
\item[Conform]{(high-conformity)}.
\end{description}

@@

\subsection{Interlocks and Solution Quality}
Figure \ref{fig:quality} shows the distribution of solution quality over
the 1000 simulation runs.
The quality shown is averaged over the ending solutions of all 100 agents.
For the low-conformity best-neighbor strategy,
all network topologies perform comparably.
For the conform strategy however,
both the small-world and preferential attachment networks
perform notably worse, while the random-pod network actually performs
very slightly better.
It is not surprising that the small-world and preferential attachment networks
show reduced solution quality in a high-conformity settings.
In such settings, it is well-known that information cascades can lead to the adoption of initially popular but sub-optimal solutions
\cite{banerjee_simple_1992}.
So it is remarkable that the random-pod network does not show this behavior.

\begin{figure}
\begin{center}
\includegraphics[width=5in]{fig-netdelib-results-net-boxplot.png}
\caption{TODO}
\label{fig:quality}
\begin{description}
\end{description}
\end{center}
\end{figure}

For more insight into the mechanism behind the random-pod network's
performance in the high-conformity case,
we examine the progress of the agents over the course of the simulation.
In particular, we focus on the basins of attraction associated with each agent's solution.
The basins of attraction are sets of similar solutions that all converge
to the same local maximum under a simple hill-climbing algorithm.

Figure \ref{fig:transitions} shows, on average,
how many agents switch basins of attraction at each timestep.
We further divide these transitions by whether the new basin's maximum is
higher or lower than the previous basin's.
We observe a qualitative difference between the behavior of the agents in the
random-pod network as compared to the other two networks.
In the small-world and preferential attachment networks,
the number of transitions decreases quickly, with the number of positive
transitions slightly larger than the number of negative.
For random-pod however,
the number of positive transitions remains notably higher than the number
of negative transitions.
While agents in the random-pod network are just as likely to adopt
a popular sub-optimal solution temporarily,
they more frequently continue trying new solutions until finding an
improvment.

\begin{figure}
\begin{center}
\includegraphics[width=5in]{fig-conform-transition-basin.png}
\caption{TODO}
\label{fig:transitions}
\begin{description}
\end{description}
\end{center}
\end{figure}

\subsection{Interlocks and Convergence Speed}
Figure \ref{fig:convergence} shows the distribution of time steps necessary for agents to converge to a stable solution.
In the low-conformity best-neighbor case,
random-pod performs comparably to preferential attachment,
both of which converge slightly faster than the small-world network.
All networks converge considerably slower in the high-conformity case.
While improved performance might be expected to come at the cost of slower
speed,
the random-pod network's median convergence is actually faster than that of the other two networks.
However, the random-pod networks slowest runs are notably slower than those for the other networks.

\begin{figure}
\begin{center}
\includegraphics[width=5in]{fig-netdelib-converge-net-boxplot.png}
\caption{TODO}
\label{fig:convergence}
\begin{description}
\end{description}
\end{center}
\end{figure}

\subsection{Interlocks and Polarization}
Figure \ref{fig:basins} shows the number of basins occupied at a given time
step, averaged over all 1000 runs of the conform strategy.
While the small-world and preferential attachment networks quickly begin to
level off and plateau at about 20 basins,
the random-pod network typically continues to converge until all agents have reached consensus on a single basin.
The small-world and preferential attachment networks thus exhibit polarization
(specifically multi-polar polarization or Balkanization),
while the random-pod network does not.
As we saw earlier in Figure \ref{fig:transitions},
the small-world and preferential attachment networks lead to a fast convergence to popular but sub-optimal solutions.
The observed polarization can be attributed to different neighborhoods within these networks converging to different sub-optimal solutions.
Thus the same mechanism that allows the random-pod network to prevent information cascades allows it to prevent polarization.

\begin{figure}
\begin{center}
\includegraphics[width=5in]{fig-conform-basin.png}
\caption{TODO}
\label{fig:basins}
\begin{description}
\end{description}
\end{center}
\end{figure}

\section{Discussion}
We have proposed and evaluated an empirically-motivated, agent-based model of social learing,
based on the interlocking network structure observed across many successful large-scale participatory projects.
Our central finding suggests that interlocks can reduce the negative effects of conformity and social influence in social-learning,
and thereby improve deliberation and participatory governance.

Groups can exhibit conformity for a number of reasons.
Conformity can stem from social factors such as trust or social loafing.
Alternatively, conformity can supplement individual knowledge and abilities.
In the low-conformity best-neighbor strategy, we do not observe a substantial difference in output quality between the interlock network and other networks. In all cases we have looked at, interlock networks perform comparably to,
or better than other common social networks. 
Our results suggest that interlocking network structure may contribute to the success of large-scale participatory projects, and may be a useful tool for the design of large-scale sociotechnical systems.

We attribute the protective effects of interlocking network structure to its ability to stop information cascades.
A closer look at the individual agents in our model shows that conformity lowers solution quality by trapping agents in separate sub-optimal yet popular solutions and preventing them from exploring alternatives.
This outcome is consistent with the well-known phenomenon of information cascades.
In interlock networks however, a sub-optimal solution might become popular in one pod,
but members of that pod are also subject to social influence from other pods.
Those other pods are unlikely to have adopted the same sub-optimal solution,
preventing it from spreading outside its initial pod and potentially even exerting enough social influence to replace it with an improved alternative.
Similarly, by providing individuals with diverse social influence from throughout the wider network, our findings suggest interlocks can reduce polarization.

\subsection{Limitations and Future Work}
Our work is limited by several assumptions, which require further investigation to achieve greater generalizability.
We make simplifying assumptions including: a static objective function, a homogeneous strategy across all agents, and truthful agents.
All of these assumptions can be violated in some real-world settings.
We have also assumed that agents have access to the full solution string, and that agents incorporate all bits of the string into their quality estimates.
Extremely complex real-world tasks might be better-modeled by a collection of sub-tasks, with agents having access to only a subset of the solution string.
We have also limited our study to learning strategies that explore new solutions solely through small mutations of existing solutions.
Other potential strategies might generate entirely new solutions, for example, by combining parts of multiple previous solutions.
Such generative strategies would add the ability to explore a wider range of solutions.

We also note that random assignment is just one potential way to assign pods.
Other assignment methods may create interlock networks with different properites,
such as longer mean shortest path lengths.
In fact, we have explored some other methods beyond the scope of this paper,
finding that they do alter some properties of the resulting interlock network,
but are nonetheless consistent with our present findings.

Finally, as the result of a simulation model, our findings require empirical validation.
We have conducted some initial lab experiments and interviews, which appear consistent with our findings, but further work is needed.

\subsection{Conclusion}
Our findings suggest that interlock networks, composed of small pods linked by common membership, contribute to the success of large-scale social learning and participatory decision-making, particularly when the topic is complex and social influence is strong. They do so by preventing information cascades and polarization.
These findings have implications for a range of contexts: participatory government and budgeting, worker-owned cooperatives, grassroots social movements, and the governance of decentralized technology. By enabling better and more participatory decision-making at larger scales, we hope this work will contribute to democratizing the governance of large sociotechnical systems, and empowering the individuals impacted by those systems.

%%
%% The acknowledgments section is defined using the "acks" environment
%% (and NOT an unnumbered section). This ensures the proper
%% identification of the section in the article metadata, and the
%% consistent spelling of the heading.
\begin{acks}
[redacted for peer review]
\end{acks}

@@

\section{Online Resources}
The code and data used in this paper are available online at [redacted for peer review].
